---
title: "Functions in ctdEvaluator Package"
author: Jinglan Qiu, Weiliang Qiu
date: "`r Sys.Date()`"
output:
  html_document:
    number_sections: true
    toc: true
header-includes:
- \usepackage{setspace}
- \doublespacing
vignette: |
  %\VignetteIndexEntry{Functions in ctdEvaluator Package} 
  %\VignetteKeyword{ctdEvaluator} 
  %\VignetteEngine{knitr::rmarkdown} 
  \usepackage[utf8](inputenc)
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.width = 12, 
  fig.height = 8,
  fig.align = "center"
)
```


# Introduction

There are abundant Bulk RNAseq datasets publicly available, while single cell RNAseq (scRNAseq) datasets are still expensive to obtain. One limitation of Bulk RNAseq datasets is that its assumption (i.e., all cells in a tissue have the same gene expression) is usually violated. This would result in misleading results if we do not handle potential confounding effect of cell type before down-streaming analysis, such as differential gene expression analysis. To utilize bulk RNAseq datasets, several cell-type deconvolution methods have been proposed to restore cell type information for bulk RNAseq datasets. However, it is challenging to evaluate the estimated cell type proportions for a bulk RNAseq data with unknown cell-type information. This R package *ctdEvaluator* contains functions to address this challenging question.

We implemented our novel evaluation methods in the current version of *ctdEvaluator*, which contains 2 functions *MuSiCEvaluator2scRNAseq* (input 2 scRNAseq datasets: one scRNAseq dataset as reference; the other scRNAseq dataset used to construct bulk RNAseq with known cell type proportions) and *MuSiCEvaluator* (input 1 scRNAseq dataset as reference and one bulk RNAseq dataset whose cell type proportions to be estimated). 
The function *MuSiCEvaluator2scRNAseq* can be used to evaluate performance of MuSiC cell type deconvolution. The function 
*MuSiCEvaluator* can be used to estimate cell type proportions in real applications.

# Dependent R/Bioconductor packages

*ctdEvaluator* depends on R package *MuSiC*, *iPlotFun*, and Bioconductor package *Biobase*.

*MuSiC* R package has not been published on CRAN or Bioconductor yet. You can install it via Wang et al's instruction in their github website https://github.com/xuranw/MuSiC and their tutorial website https://xuranw.github.io/MuSiC/articles/MuSiC.html, which is copied below:
```{r eval = F, echo = T, message = F}

# install devtools if necessary
install.packages('devtools')

# install the MuSiC package
devtools::install_github('xuranw/MuSiC')

# load
library(MuSiC)

```


*iPlotFun* is an R package developed by Weiliang Qiu and Luc Essermeant to facilitate data visualization for NCES (non-clinical efficacy & safety) projects. Please contact Weiliang Qiu (Weiliang.Qiu@sanofi.com) to get a copy of *iPlotFun* R package.


To install *Biobase*, please use the following commands based on the website https://bioconductor.org/packages/release/bioc/html/Biobase.html:
```{r eval = F, echo = T, message = F}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("Biobase")

```

# Datasets

*ctdEvaluator* package contains 4 datasets: 3 scRNAseq datasets and 1 bulk RNAseq dataset.

## scRNAseq: esRef.EMTAB

The dataset *esRef.EMTAB.sce* is an SingleCellExperiment object
storing a scRNAseq dataset that can be used as a reference scRNAseq in MuSiC algorithm. It contains 25453 genes and 748 samples (i.e., cells).
The 748 cells belong to 6 non-T2D subjects.
Among the 748 cells, there are 4 cell types: alpha ($n=443$ cells), beta ($n=171$ cells), delta ($n=59$ cells), and gamma ($n=75$ cells). This dataset is a subset of EMTAB scRNAseq dataset listed in Wang et al.'s tutorial website https://xuranw.github.io/MuSiC/articles/MuSiC.html.

To load *esRef.EMTAB.sce*, we can use the following R code:
```{r eval = T, echo = T, message = F}
# load R package ctdEvaluator
library(ctdEvaluator)
library(SummarizedExperiment)
library(SingleCellExperiment)

# load data
data(esRef.EMTAB.sce)

print(esRef.EMTAB.sce)

# unique subject ids
print(unique(esRef.EMTAB.sce$SubjectName))

# frequencies of cell types
print(table(esRef.EMTAB.sce$cellType))

```

## scRNAseq: esRef.EMTAB.fakeCT

The dataset *esRef.EMTAB.fakeCT.sce* is an SingleCellExperiment object
storing a scRNAseq dataset that can be used as a reference scRNAseq in MuSiC algorithm. It contains 25453 genes and 299 samples (i.e., cells).
The 299 cells belong to 6 non-T2D subjects.
Among the 748 cells, there are 4 cell types: acinar ($n=112$ cells), ductal ($n=135$ cells), PSC ($n=23$ cells), and unclassified endocrine ($n=29$ cells). 

We relabeled the cell types so that we can use it as a reference scRNAseq to estimate cell types (alpha, beta, delta, and gamma) using MuSiC and expect MuSiC could not perform well for this scenario.

\begin{tabular}{l|l}
\hline
true cell type & relabeled cell type\\
\hline
acinar & gamma\\
ductal & delta \\
PSC & alpha\\
unclassified endocrine & beta\\
\hline
\end{tabular}

To load *esRef.EMTAB.fakeCT*, we can use the following R code:
```{r eval = T, echo = T, message = F}
# load data
data(esRef.EMTAB.fakeCT.sce)

print(esRef.EMTAB.fakeCT.sce)

# unique subject ids
print(unique(esRef.EMTAB.fakeCT.sce$SubjectName))

# frequencies of true cell types
print(table(as.character(esRef.EMTAB.fakeCT.sce$cellType.true)))

# frequencies of relabeled cell types
print(table(esRef.EMTAB.fakeCT.sce$cellType))

# cross table

print(table(as.character(esRef.EMTAB.fakeCT.sce$cellType.true), esRef.EMTAB.fakeCT.sce$cellType))


```

## scRNAseq: es4Bulk.XinT2D

The dataset *es4Bulk.XinT2D* is an SingleCellExperiment object
storing a scRNAseq dataset that can be used to construct Bulk RNAseq with known cell type proportions. It contains 39849 genes and 1492 samples (i.e., cells).
The 1492 cells belong to 6 T2D subjects and 12 non-T2D subjects.
Among the 1492 cells, there are 4 cell types: alpha ($n=886$ cells), beta ($n=472$ cells), delta ($n=49$ cells), and gamma ($n=85$ cells). This dataset is listed in Wang et al.'s tutorial website https://xuranw.github.io/MuSiC/articles/MuSiC.html.

To load *es4Bulk.XinT2D*, we can use the following R code:
```{r eval = T, echo = T, message = F}

# load data
data(es4Bulk.XinT2D.sce)

print(es4Bulk.XinT2D.sce)

# unique subject ids
print(unique(es4Bulk.XinT2D.sce$SubjectName))

# frequencies of cell types
print(table(es4Bulk.XinT2D.sce$cellType))

```


## bulk RNAseq: esBulk.XinT2D

The dataset *esBulk.XinT2D* is an SingleCellExperiment object
storing a Bulk RNAseq dataset that was constructed by the scRNAseq dataset *es4Bulk.XinT2D* by averaging gene expression across all cells for each subject. The real cell type proportions are hence known for *esBulk.XinT2D*. It contains 39849 genes and 18 samples (i.e., subjects).

To load *esBulk.XinT2D*, we can use the following R code:
```{r eval = T, echo = T, message = F}

# load data
data(esBulk.XinT2D.sce)

print(esBulk.XinT2D.sce)

# unique subject ids
print(unique(esBulk.XinT2D.sce$SubjectName))

# true cell type proportions
colDat = colData(esBulk.XinT2D.sce)
print(dim(colDat))
print(colDat)


```



# Examples


## esBulk.XinT2D.sce as bulk RNAseq and esRef.EMTAB as reference

The function *MuSiCEvaluator* is to evaluate estimated cell type proportions for Bulk RNAseq data in real applications, in which cell type info for the Bulk RNAseq data is unknown.

We use *esRef.EMTAB.sce* scRNAseq data as reference and use *esBulk.XinT2D.sce* as bulk RNAseq data. Note that *esBulk.XinT2D.sce* is constructed based on the scRNAseq data *es4Bulk.XinT2D.sce*. Hence, we expect high-value and consistent pseudo $R^2$ based on results in previous section (*esRef.EMTAB.sce* as reference and *es4Bulk.XinT2D.sce* used to construct bulk RNAseq).


```{r message = F, eval = T, echo = T, warning = F}
#####################

# load reference scRNAseq data
data(esBulk.XinT2D.sce)
print(esBulk.XinT2D.sce)

# First do cell type deconvolution, then evaluate the estimates.
# For testing purpose in this vignettes, 
#  we set number of bootstrapping 'nBoot=3'.
# In real applications, please set nBoot >= 20.
# Since it is computationally intensive, it would take a long running time 
#   when 'nBoot' is too large.
nBoot = 3
nPerm = NULL
res = MuSiCEvaluator(esBulk = esBulk.XinT2D.sce, esRef = esRef.EMTAB.sce, 
                              sid = "SubjectName", 
                              cellType = "cellType", 
                              nBoot = nBoot, 
                              seed = 1234567,
                              nPerm = nPerm)
print(names(res))

```

The plot of pseudo $R^2$ across different reference scRNAseq datasets for each subject is shown below.
```{r message = F, eval = T, echo = T, warning = F}

##############
# trajectories of pseudo R2 across different reference scRNAseq datasets;
# cell type proportions estimated via MuSiC 
# note that pseudo R2 can be negative
R2.MuSiC.boot.LF3 = res$R2.MuSiCFrame.boot
order.dataset = c("orig", paste("boot", 1:nBoot, sep=""))

g = iPlotFun::trajPlot1Grp(dat = R2.MuSiC.boot.LF3, 
         time = "grp", 
         y = "R2.MuSiC", 
         sid = "sid", 
         numericTime = FALSE,
         order.time = order.dataset, 
         xlab = "", 
         ylab = "pseudo R2", 
         title = "MuSiC: esBulk.XinT2D.sce and esRef.EMTAB.sce", 
         xAngle = 45
)
print(g)
```

Similarly, the plot of $R^2$ across different reference scRNAseq datasets for each subject is shown below.
```{r message = F, eval = T, echo = T, warning = F}
# trajectories of R2 across different reference scRNAseq datasets;
# cell type proportions estimated via NNLS 
R2.NNLS.boot.LF3=res$R2.NNLSFrame.boot

g2 = iPlotFun::trajPlot1Grp(dat = R2.NNLS.boot.LF3, 
         time = "grp", 
         y = "R2.NNLS", 
         sid = "sid", 
         numericTime = FALSE,
         order.time = order.dataset, 
         xlab = "", 
         ylab = "R2", 
         title = "NNLS: esBulk.XinT2D.sce and esRef.EMTAB.sce", 
         xAngle = 45
)
print(g2)

```

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

# References

Wang, X., Park, J., Susztak, K. et al. Bulk tissue cell type deconvolution with multi-subject single-cell expression reference. Nat Commun 10, 380 (2019). https://doi.org/10.1038/s41467-018-08023-x

